Individual foraging specialisation and success change across experience in a virtual predator-prey system

### 2025-04-09 manuscript/manuscript.Rmd
On Windows --> discussion repassed. fig2 updt

---
title: "Individual foraging specialization and success change across experience in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: ecology.csl
---

@@

The capacity of predators to match their hunting tactics to their prey and to optimize their skills at using them over successive encounters is hypothesized to promote individual foraging specialisation. Through this process, predators should be able to capture more prey. Yet, there are limited empirical assessments showing links between past experience, foraging specialization, and hunting success at the individual level, due to the challenges of monitoring direct interactions in the wild. In this study, we used a virtual predator-prey system (the game *Dead by Daylight*) to investigate how individual predator foraging specialization and success developed across repeated interactions with prey. We found that 68% of predators became either increasingly specialized by always moving at a fast pace (specialised cursorial tactic), or generalists by transitioning between slow and fast speeds (generalist ambush tactic). The predators' strategies were partially matched to their prey's speed, suggesting that changes in hunting behaviour were driven by repeated encounters with their prey. Specialist and generalist foragers achieved similar success overall, although specialist hunters had a slight advantage. Hence, our findings suggest that experience may promote behavioural diversification in predator-prey systems.

@@

Predators express a diverse array of foraging tactics to capture their prey. In natural populations, individuals specialise in the use of tactics such as ambushing and roaming, or switch tactics based on the prey that they encounter, often resulting in differences in prey consumption [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. The success of specialist and generalist strategies depend on ecological factors: when resources are abundant, specialist hunters achieve greater success, while generalist hunters prevail under fluctuating conditions [@Terraube2011;@Terraube2014;@phillips.etal2017]. While evolutionary history can partly explain how specialist and generalist species will fare under given ecological conditions, learning also plays an important part in shaping predator foraging behaviour [@wooster.etal2023]. This has prompted ecologists to hypothesize that experience may underly associations between foraging specialization and prey capture as predators learn and become efficient hunters through repeated encounters with their prey [@estes.etal2003;@woo.etal2008;@phillips.etal2017]. Yet, it remains challenging to monitor individual predator experience, limiting our ability to uncover how predator foraging behaviour shapes prey capture over time.

Recent evidence suggests that the resource fluctuations predators experience over time (e.g., temporal changes in prey behaviour) may be key in shaping individual behavioural specialization and its fitness consequences [@patrick.weimerskirch2014b;@vandenbosch.etal2019;@santoro.etal2019;@manlick.etal2021]. Theory outlines two contrasting scenarios, with outcomes defined by the predator's ability to learn the optimal strategy in response to resource variability. In the first scenario, individuals should gradually specialize in a foraging tactic (e.g., always moving fast) if it enables them to consistently achieve successful prey captures while learning to hunt in environments with predictable (or stable) resources [@tinker.etal2009;@dukas2019]. Such conditions favor specialist hunters, as they require less energy and time to search for, capture, and handle prey [@weimerskirch2007;@woo.etal2008;@potier.etal2015]. For example, habitat specialization under stable resources is associated with reduced foraging effort and increased offspring growth rates in Herring Gulls (*Larus argentatus*) [@vandenbosch.etal2019]. In the second scenario, predators are expected to be more generalist by learning to use a variety of hunting tactics when they hunt in highly variable environments. Indeed, while there can be costs to attempt novel tactics by trial and error, predators can reduce the consequences of uncertainty in environments under fluctuating resources by learning to adjust their tactic to the prey that they encounter [@dall2010;@mathot.etal2012;@stephens1993;@ishii.shimada2010;@kelley.magurran2011]. @szopa-comley.ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Under these two scenarios, if predators can learn to adjust their tactics to match the prey fluctuation regime they experience, both specialist and generalist hunters should achieve similar hunting success and coexist [@phillips.etal2017].

In this study, we use data from players in the online videogame *Dead by Daylight* to investigate 1) how repeated interactions with prey shape predator foraging specialization and 2) how the relationship between specialization and foraging success changes over time. *Dead by Daylight* simulates a predator-prey interaction across complex virtual environments, providing a system to test ecological hypotheses on foraging behaviour, group selection, and social interactions [reviewed in @montiglio2025]. In this game, four prey players forage for resources while avoiding predation by a fifth player. The predator population is composed of individuals that vary consistently in their use of hunting tactics along an ambush-cursorial continuum: at one extreme hunters move more slowly and ambush prey; at the other extreme, hunters roam the virtual environment at high velocities [@fraserfranco.etal2022].

```{r fig1, out.width = "100%", fig.cap = "**Figure 1**. Hypothesis describing how prey-mediated individual foraging specialization emerges over repeated interactions. Panels A and B display two scenarios with different outcomes at the population level, but similar outcomes at the individual level. A) all predators encounter prey with similar attributes (e.g., some encountered prey that were always fast while others encountered slower prey more often) through time. They differ in their mean behaviour (tactic) which is adjusted to the prey that they encountered. Because encounters where predictable, they all display similar degrees of individual foraging specialization, resulting in the population being specialized. B) all predators encounter varying types of prey (e.g., in some encounters the prey were fast, in others the prey were slower), such that they also display similar degrees of individual foraging specialization as in panel A. However, in this case, they all become generalist hunters, resulting in the population distribution being wider. C) A scenario where predators display varying degrees of foraging specialization. In this case, some individuals consistently encountered similar prey, and thus specialized, while others encountered varying groups of prey, thus becoming generalists. Ultimately, those that learn to specialise on the proper tactic given prey encounters have greater success than those that didn't, resulting in no differences between specialists and generalists."}

@@

To test the hypothesis that repeated interactions with prey shape predator foraging specialization, we analyze the movement speed of predators and their prey over time. To assess how foraging specialization emerges throughout experience, we define the level of individual foraging specialization as the intra-individual variance (IIV) in movement speed across matches following the definition by @cleasby.etal2015. Low IIV describes individual foraging specialization (i.e., always using an ambush or cursorial tactic) whereas high IIV describes a flexible use of foraging tactics (i.e., generalist predators switching between ambush and cursorial tactics, see Methods for details). First, we expect predators to differ in the use of their hunting tactic (i.e., mean movement speed) which will be tuned to the mean speed of the prey that they encounter (Figure 1A-B). Second, we predict that if all predators consistently encounter prey moving at similar speeds, then the population as a whole (i.e., all individuals) should specialize on moving at similar speeds, resulting in a narrow behavioural niche (Figure 1A). In contrast, if all predators encounter groups of prey moving at varying speeds, then they should all become more flexible in their tactic use (i.e., generalists), adjusting their speed to the prey from one encounter to the other, resulting in a higher population variance in foraging behaviour (Figure 1B). In both scenarios, differences among individuals in IIV across experience should be low (i.e., similar individual foraging specialization, Figure 1A-B). Alternatively, differences among individuals in foraging specialization may emerge if they experience different interactions with their prey. In this case, predators that encountered prey moving at similar speeds across experience should specialize on moving at a more fixed speed, while predators that encountered different prey speeds across matches should adopt a generalist movement strategy, resulting in an increase in among individual differences in IIV (Figure 1C). Lastly, if we detect such prey-dependent fine-tuning with experience, then specialist and generalist hunters should attain equal success, resulting in no apparent relationship between foraging specialization and prey capture (Figure 1C).

@@

 We used a multivariate double-hierarchical generalized linear model (MDHGLM) to quantify a) changes in the average foraging tactic (i.e., the mean movement speed) and in the degree of specialization (i.e., IIV in movement speed) over successive matches for each player; and b) the relationship between the average tactic and specialization, and hunting success [@lee.nelder2006;@cleasby.etal2015;@mitchell.etal2016]. For each experience level, we fitted the model on three traits: the predator's speed ($Y_{1}$) and the prey's speed ($Y_{2}$) using a Gaussian error distribution, and the predator's hunting success ($Y_{3}$) using a beta-binomial error distribution. Fixed and random effect structure was trait-specific and the same for all experience levels. For the predator and the prey speed, we modeled the mean and the residual standard deviation (eqn 1.1, 1.2 and 2.1, 2.2). For both traits, we controlled for the standardized prey rank on the mean and the residual standard deviation. We included game environment, predator avatar, and individual predator identity random effects on the mean part of the model. The residual standard deviation included individual predator identity random effects exclusively. For hunting success, we modeled the mean only (eqn 3) and controlled for the standardized prey rank and game duration, and included an individual identity random effect.

The model simultaneously estimated five among-individual variance components with their covariances for each experience level, resulting in a 15x15 (co)variance matrix. For each experience level, the model estimated among individual variance in mean predator speed, mean prey speed, and mean hunting success. This allows us to test for individual differences in tactic use (i.e., along the ambush-cursorial continuum), in the mean speed of the prey encountered, and in mean prey consumption. The model also estimated among individual variance in intra-individual variation (IIV) for both prey and predator speed. This allows us to test if predators do not experience the same degree of prey heterogeneity and if they differ in individual specialization. Finally, the model estimated all pairwise covariances at the individual level among traits, among experience, and across traits and experience. The resulting 15x15 (co)variance matrix thus provides the structure of the predator-prey trait interactions as well as the relationship between specialization and success across experience at the individual-level.

@@

At the population level, the mean predator speed remained stable as predators progressed from novices to advanced hunters (Figure 2A and Appendix 1: Table S1). Likewise, the variation in speed did not change with experience, indicating that foraging specialization remained stable at the population level (Figure 2A and Appendix 1: Table S1). At the individual level, differences among predators in their mean speed were weak and did not change across experience (Figure 2A and Appendix 1: Table S2). However, predators displayed minor differences in foraging specialization as novices, with differences increasing slightly as individuals gained experience (Figure 2A and Appendix 1: Table S2). `r toString(scales::percent(percent_tab2[1,]$percentage, accuracy = 1))` of the population switched from a flexible to a more specialized hunting tactic, and vice-versa, as predators gained experience (i.e., >0.2 change in standard deviation; Figure 3). In contrast, `r toString(scales::percent(percent_tab2[2,]$percentage, accuracy = 1))` displayed lower changes (i.e., >0.05 and <0.2 change in standard deviation) and `r toString(scales::percent(percent_tab2[3,]$percentage, accuracy = 1))` showed almost no change with experience (i.e., <0.05 change in standard deviation).

@@

```{r fig3, out.width = "100%", fig.cap = "**Figure 3**. Individual behavioural distribution of the predators' speed drawn from the MDHGLM estimates for individuals that had (A) the greatest increase and (B) the greatest decrease in specialisation. The background (gray) distributions are for when individuals were novice hunters, and the foreground (blue) distributions with solid contour lines are for when they were advanced hunters. Both figure panels are ordered by ascending degree of change in specialization. We built the figure by first subtracting the estimated standard deviation of all individuals as novices with their standard deviation as advanced hunters, and then selected those with the greatest difference using the 25% and 75% quantiles. Individuals on panel A have an increase in specialization equal or greater than 0.2 standard deviations, while those on panel B have an increase in flexibility equal to or greater than 0.2 standard deviations."}

@@

Predators that were on average faster had lower IIV in speed (Figure 4A-B). Thus, faster predators specialized to a higher degree than slower predators. As predators gained experience, these strategies were increasingly defined as the correlation changed from -0.62 (-0.70, -0.55 $95\%\ CI$) among novices to -0.78 (-0.83, -0.72 $95\%\ CI$) among advanced hunters (Figure 4C). Faster predators tended to encounter faster prey; they also encountered groups of prey groups of prey that were more similar from match to match (Figure 4A-B), irrespective of their experience (Figure 4C).

There was no correlation (-0.02, -0.14, 0.01 $95\%\ CI$) between mean prey speed and IIV in predator speed when predators were novice (Figure 4A), but the correlation increased to -0.16 (-0.27, -0.04 $95\%\ CI$) at the advanced stage (Figure 4B), such that advanced hunters that encountered faster prey tended to be more specialized (Figure 4C). Predators with a larger IIV in speed encountered a greater range of prey speeds, but this effect was weak (Figure 4A-B). Hence, predators that encountered groups of prey with similar speeds tended to be more specialized in their tactic, while those that encountered groups of prey with varying speeds were less specialised. However, this correlation did not change with experience (Figure 4C).

@@

## Success along the foraging specialization continuum

@@

A general assumption of predator-prey studies is that predators maximize success by matching their tactic to their prey [@abrams2000], yet, it has remained largely unknown whether this results from predators learning how to hunt their prey in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found that while predators in *Dead by Daylight* did not show any increase in either behavioural specialization or flexibility with experience at the population level, individual predators developed their own behavioural trajectories over successive encounters with the prey. This was characterized by a continuum of hunting styles ranging from a slower and flexible strategy to a highly specialized fast-paced strategy, with some predators switching between the two across experience. The behavioural strategies of predators and prey tended to match across experience, suggesting that both were adjusting to each other. Specialized cursorial hunters encountered similar groups of prey, while slower and flexible hunters encountered varying groups of prey. Overall, predators along the flexible-specialist continuum achieved similar success, although flexible hunters were slightly less successful as advanced hunters.

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements, while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@endler1991;@bro-jorgensen2013;@wilson.etal2018;@szopa-comley.ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they encountered faster prey more often. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@kishida.etal2006;@kishida.etal2009;@edgell.rochette2009;@mcghee.etal2013]. Indeed, 56% of the predator population altered its behaviour with experience to different degrees, including 12% that even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adjusting to each other, similar to Red Queen dynamics [@brockhurst2014].

Classical theory predicts that individual foraging specialization should maximize prey consumption by optimizing foraging efficiency [@stephens1986], but other work suggests that flexibility and specialization could both be adaptive as resources fluctuate [@woo.etal2008;@phillips.etal2017]. Our results support empirical findings indicating that flexible and specialist foragers obtained similar success [@weimerskirch2007;@woo.etal2008;@potier.etal2015;@phillips.etal2017;@courbin.etal2018], although flexible foragers were slightly less successful. Both strategies appeared to emerge in part from individuals learning how to hunt their prey, resulting in a general increase in success in the population. However, there were still considerable differences in success among individuals through time, suggesting that some predators were limited in their capacity to match their tactic to their prey or to increase their success through other means.

Specialist foragers were faster and probably better equipped to hunt the faster and more difficult prey in *DBD*. However, if the prey responded to fast predators by also being faster, then hunting at high speeds resulted in challenging encounters for these predators too, thereby decreasing the benefits of using this tactic (Figure 3). Thus, specializing probably compensated for the difficulty of hunting prey at high speeds by helping predators to better predict the location and movement of their prey. On the other hand, flexible foragers encountered a larger range of prey moving at slower speeds. Yet, because the prey increased their speed with experience, the benefits of being able to hunt multiple prey types for flexible hunters may have come at the cost of not being adept at capturing faster prey [@pintor.etal2014]. Such trade-offs may reflect limitations in learning all the skills required to successfully hunt all types of prey [@healy1992;@belisle1997;@dukas2019], particularly if the skills required to hunt slower prey are nontransferable to faster prey. As the benefits of each hunting style changed under different scenarios, the combination of predatory styles in this virtual system reflects how resource fluctuations can maintain fitness equilibrium within populations, resulting in behavioural diversity in predator-prey systems [@tinker.etal2008;@woo.etal2008;@mougi.kishida2009].

@@

We found support for our prediction that individual foraging specialization changed across experience and predator-prey interactions. Our results suggest that predators learned with experience, as their success increased and their speed remained matched to the speed of their prey.  Even though individuals were not all equally successful, both specialized and flexible hunters achieved similar success overall. A potential caveat is that the more flexible hunters might have experimented with various tactics out of boredom, which could impede ecologically realistic interpretations of our data. However, the consistent association between this tactic and heterogeneous prey groups gives us confidence that it emerged from their interactions. One limitation of our study was that we couldn't monitor all the matches of the prey, which prevented us from assessing their responses to the predator through their experience. Hence, future studies should aim at monitoring reciprocal behavioural dynamics through time, which may reveal important insights on the mechanisms driving predator-prey systems.

>>>>

---
title: "Individual foraging specialisation and success change across experience in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: ecology.csl
---

@@

The capacity of predators to match their hunting tactics to their prey and to optimise their skills at using them over successive encounters is hypothesised to promote individual foraging specialisation. Through this process, predators should be able to capture more prey. Yet, there are limited empirical assessments showing links between past experience, foraging specialisation, and hunting success at the individual level, due to the challenges of monitoring direct interactions in the wild. In this study, we used a virtual predator-prey system (the game *Dead by Daylight*) to investigate how individual predator foraging specialisation and success developed across repeated interactions with prey. We found that 68% of predators became either increasingly specialised by always moving at a fast pace (specialised cursorial tactic), or generalists by transitioning between slow and fast speeds (generalist ambush tactic). The predators' strategies were partially matched to their prey's speed, suggesting that changes in hunting behaviour were driven by repeated encounters with their prey. Specialist and generalist foragers achieved similar success overall, although specialist hunters had a slight advantage. Hence, our findings suggest that experience may promote behavioural diversification in predator-prey systems.

@@

Predators express a diverse array of foraging tactics to capture their prey. In natural populations, individuals specialise in the use of tactics such as ambushing and roaming, or switch tactics based on the prey that they encounter, often resulting in differences in prey consumption [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. The success of specialist and generalist strategies depend on ecological factors: when resources are abundant, specialist hunters achieve greater success, while generalist hunters prevail under fluctuating conditions [@Terraube2011;@Terraube2014;@phillips.etal2017]. While evolutionary history can partly explain how specialist and generalist species will fare under given ecological conditions, learning also plays an important part in shaping predator foraging behaviour [@wooster.etal2023]. This has prompted ecologists to hypothesise that experience may underly associations between foraging specialisation and prey capture as predators learn and become efficient hunters through repeated encounters with their prey [@estes.etal2003;@woo.etal2008;@phillips.etal2017]. Yet, it remains challenging to monitor individual predator experience, limiting our ability to uncover how predator foraging behaviour shapes prey capture over time.

Recent evidence suggests that the resource fluctuations predators experience over time (e.g., temporal changes in prey behaviour) may be key in shaping individual behavioural specialisation and its fitness consequences [@patrick.weimerskirch2014b;@vandenbosch.etal2019;@santoro.etal2019;@manlick.etal2021]. Theory outlines two contrasting scenarios, with outcomes defined by the predator's ability to learn the optimal strategy in response to resource variability. In the first scenario, individuals should gradually specialise in a foraging tactic (e.g., always moving fast) if it enables them to consistently achieve successful prey captures while learning to hunt in environments with predictable (or stable) resources [@tinker.etal2009;@dukas2019]. Such conditions favor specialist hunters, as they require less energy and time to search for, capture, and handle prey [@weimerskirch2007;@woo.etal2008;@potier.etal2015]. For example, habitat specialisation under stable resources is associated with reduced foraging effort and increased offspring growth rates in Herring Gulls (*Larus argentatus*) [@vandenbosch.etal2019]. In the second scenario, predators are expected to be more generalist by learning to use a variety of hunting tactics when they hunt in highly variable environments. Indeed, while there can be costs to attempt novel tactics by trial and error, predators can reduce the consequences of uncertainty in environments under fluctuating resources by learning to adjust their tactic to the prey that they encounter [@dall2010;@mathot.etal2012;@stephens1993;@ishii.shimada2010;@kelley.magurran2011]. @szopa-comley.ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Under these two scenarios, if predators can learn to adjust their tactics to match the prey fluctuation regime they experience, both specialist and generalist hunters should achieve similar hunting success and coexist [@phillips.etal2017].

In this study, we use data from players in the online videogame *Dead by Daylight* to investigate 1) how repeated interactions with prey shape predator foraging specialisation and 2) how the relationship between specialisation and foraging success changes over time. *Dead by Daylight* simulates a predator-prey interaction across complex virtual environments, providing a system to test ecological hypotheses on foraging behaviour, group selection, and social interactions [reviewed in @montiglio2025]. In this game, four prey players forage for resources while avoiding predation by a fifth player. The predator population is composed of individuals that vary consistently in their use of hunting tactics along an ambush-cursorial continuum: at one extreme hunters move more slowly and ambush prey; at the other extreme, hunters roam the virtual environment at high velocities [@fraserfranco.etal2022].

```{r fig1, out.width = "100%", fig.cap = "**Figure 1**. Hypothesis describing how prey-mediated individual foraging specialisation emerges over repeated interactions. Panels A and B display two scenarios with different outcomes at the population level, but similar outcomes at the individual level. A) all predators encounter prey with similar attributes (e.g., some encountered prey that were always fast while others encountered slower prey more often) through time. They differ in their mean behaviour (tactic) which is adjusted to the prey that they encountered. Because encounters where predictable, they all display similar degrees of individual foraging specialisation, resulting in the population being specialised. B) all predators encounter varying types of prey (e.g., in some encounters the prey were fast, in others the prey were slower), such that they also display similar degrees of individual foraging specialisation as in panel A. However, in this case, they all become generalist hunters, resulting in the population distribution being wider. C) A scenario where predators display varying degrees of foraging specialisation. In this case, some individuals consistently encountered similar prey, and thus specialised, while others encountered varying groups of prey, thus becoming generalists. Ultimately, those that learn to specialise on the proper tactic given prey encounters have greater success than those that didn't, resulting in no differences between specialists and generalists."}

@@

To test the hypothesis that repeated interactions with prey shape predator foraging specialisation, we analyze the movement speed of predators and their prey over time. To assess how foraging specialisation emerges throughout experience, we define the level of individual foraging specialisation as the intra-individual variance (IIV) in movement speed across matches following the definition by @cleasby.etal2015. Low IIV describes individual foraging specialisation (i.e., always using an ambush or cursorial tactic) whereas high IIV describes a flexible use of foraging tactics (i.e., generalist predators switching between ambush and cursorial tactics, see Methods for details). First, we expect predators to differ in the use of their hunting tactic (i.e., mean movement speed) which will be tuned to the mean speed of the prey that they encounter (Figure 1A-B). Second, we predict that if all predators consistently encounter prey moving at similar speeds, then the population as a whole (i.e., all individuals) should specialise on moving at similar speeds, resulting in a narrow behavioural niche (Figure 1A). In contrast, if all predators encounter groups of prey moving at varying speeds, then they should all become more flexible in their tactic use (i.e., generalists), adjusting their speed to the prey from one encounter to the other, resulting in a higher population variance in foraging behaviour (Figure 1B). In both scenarios, differences among individuals in IIV across experience should be low (i.e., similar individual foraging specialisation, Figure 1A-B). Alternatively, differences among individuals in foraging specialisation may emerge if they experience different interactions with their prey. In this case, predators that encountered prey moving at similar speeds across experience should specialise on moving at a more fixed speed, while predators that encountered different prey speeds across matches should adopt a generalist movement strategy, resulting in an increase in among individual differences in IIV (Figure 1C). Lastly, if we detect such prey-dependent fine-tuning with experience, then specialist and generalist hunters should attain equal success, resulting in no apparent relationship between foraging specialisation and prey capture (Figure 1C).

@@

 We used a multivariate double-hierarchical generalized linear model (MDHGLM) to quantify a) changes in the average foraging tactic (i.e., the mean movement speed) and in the degree of specialisation (i.e., IIV in movement speed) over successive matches for each player; and b) the relationship between the average tactic and specialisation, and hunting success [@lee.nelder2006;@cleasby.etal2015;@mitchell.etal2016]. For each experience level, we fitted the model on three traits: the predator's speed ($Y_{1}$) and the prey's speed ($Y_{2}$) using a Gaussian error distribution, and the predator's hunting success ($Y_{3}$) using a beta-binomial error distribution. Fixed and random effect structure was trait-specific and the same for all experience levels. For the predator and the prey speed, we modeled the mean and the residual standard deviation (eqn 1.1, 1.2 and 2.1, 2.2). For both traits, we controlled for the standardized prey rank on the mean and the residual standard deviation. We included game environment, predator avatar, and individual predator identity random effects on the mean part of the model. The residual standard deviation included individual predator identity random effects exclusively. For hunting success, we modeled the mean only (eqn 3) and controlled for the standardized prey rank and game duration, and included an individual identity random effect.

The model simultaneously estimated five among-individual variance components with their covariances for each experience level, resulting in a 15x15 (co)variance matrix. For each experience level, the model estimated among individual variance in mean predator speed, mean prey speed, and mean hunting success. This allows us to test for individual differences in tactic use (i.e., along the ambush-cursorial continuum), in the mean speed of the prey encountered, and in mean prey consumption. The model also estimated among individual variance in intra-individual variation (IIV) for both prey and predator speed. This allows us to test if predators do not experience the same degree of prey heterogeneity and if they differ in individual specialisation. Finally, the model estimated all pairwise covariances at the individual level among traits, among experience, and across traits and experience. The resulting 15x15 (co)variance matrix thus provides the structure of the predator-prey trait interactions as well as the relationship between specialisation and success across experience at the individual-level.

@@

At the population level, the mean predator speed remained stable as predators progressed from novices to advanced hunters (Figure 2A and Appendix 1: Table S1). Likewise, the variation in speed did not change with experience, indicating that foraging specialisation remained stable at the population level (Figure 2A and Appendix 1: Table S1). At the individual level, differences among predators in their mean speed were weak and did not change across experience (Figure 2A and Appendix 1: Table S2). However, predators displayed minor differences in foraging specialisation as novices, with differences increasing slightly as individuals gained experience (Figure 2A and Appendix 1: Table S2). `r toString(scales::percent(percent_tab2[1,]$percentage, accuracy = 1))` of the population switched from a flexible to a more specialised hunting tactic, and vice-versa, as predators gained experience (i.e., >0.2 change in standard deviation; Figure 3). In contrast, `r toString(scales::percent(percent_tab2[2,]$percentage, accuracy = 1))` displayed lower changes (i.e., >0.05 and <0.2 change in standard deviation) and `r toString(scales::percent(percent_tab2[3,]$percentage, accuracy = 1))` showed almost no change with experience (i.e., <0.05 change in standard deviation).

@@

```{r fig3, out.width = "100%", fig.cap = "**Figure 3**. Individual behavioural distribution of the predators' speed drawn from the MDHGLM estimates for individuals that had (A) the greatest increase and (B) the greatest decrease in specialisation. The background (gray) distributions are for when individuals were novice hunters, and the foreground (blue) distributions with solid contour lines are for when they were advanced hunters. Both figure panels are ordered by ascending degree of change in specialisation. We built the figure by first subtracting the estimated standard deviation of all individuals as novices with their standard deviation as advanced hunters, and then selected those with the greatest difference using the 25% and 75% quantiles. Individuals on panel A have an increase in specialisation equal or greater than 0.2 standard deviations, while those on panel B have an increase in flexibility equal to or greater than 0.2 standard deviations."}

@@

Predators that were on average faster had lower IIV in speed (Figure 4A-B). Thus, faster predators specialised to a higher degree than slower predators. As predators gained experience, these strategies were increasingly defined as the correlation changed from -0.62 (-0.70, -0.55 $95\%\ CI$) among novices to -0.78 (-0.83, -0.72 $95\%\ CI$) among advanced hunters (Figure 4C). Faster predators tended to encounter faster prey; they also encountered groups of prey groups of prey that were more similar from match to match (Figure 4A-B), irrespective of their experience (Figure 4C).

There was no correlation (-0.02, -0.14, 0.01 $95\%\ CI$) between mean prey speed and IIV in predator speed when predators were novice (Figure 4A), but the correlation increased to -0.16 (-0.27, -0.04 $95\%\ CI$) at the advanced stage (Figure 4B), such that advanced hunters that encountered faster prey tended to be more specialised (Figure 4C). Predators with a larger IIV in speed encountered a greater range of prey speeds, but this effect was weak (Figure 4A-B). Hence, predators that encountered groups of prey with similar speeds tended to be more specialised in their tactic, while those that encountered groups of prey with varying speeds were less specialised. However, this correlation did not change with experience (Figure 4C).

@@

## Success along the foraging specialisation continuum

@@

A general assumption in predator-prey studies is that predators maximize their success by matching their tactic to their prey [@abrams2000]. Yet ecologists have historically struggled to determine whether this results from predators learning how to hunt their prey, in part due to the challenges of studying direct interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found that individual predators in *Dead by Daylight*  developed behavioural strategies that were associated with prey behaviour over successive encounters. The strategies formed a continuum of hunting styles, ranging from a slower, generalist strategy to a highly specialised, fast-paced strategy. We also found that some predators switched strategies as they gained experience. specialised cursorial hunters encountered groups of prey moving at similar speeds, while the slower and generalist hunters encountered groups of prey with varying speeds. Overall, predators along the generalist-specialist continuum achieved similar success, although generalist hunters were slightly less successful at more advanced experience levels.

The predator population maintained its speed and degree of specialisation throughout the study period, and individuals differed only slightly in their average speeds. However, individuals displayed differences in foraging specialisation, and these differences increased slightly as individuals gained experience. Predators that hunted at higher speeds specialised in this tactic and tended to encounter groups of prey with similar speeds, while those that hunted at slower speeds were more generalists by switching tactics, and tended to encounter groups of prey moving at different speeds. On one hand, this pattern could arise if predators adjusted their strategy to their prey, because fast-paced hunting is a specialised tactic suited for prey that use rapid evasive movements, while a slower and generalist strategy can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@endler1991;@bro-jorgensen2013;@wilson.etal2018;@szopa-comley.ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialisation with experience. Since these individuals were already specialised as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they encountered faster prey more often. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@kishida.etal2006;@kishida.etal2009;@edgell.rochette2009;@mcghee.etal2013]. Indeed, 56% of the predator population altered its behaviour with experience to different degrees, including 12% that even switched from a specialist to a generalist strategy and vice-versa. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adjusting to each other, similar to Red Queen dynamics [@brockhurst2014].

Classical theory predicts that individual foraging specialisation should maximise prey consumption by optimising foraging efficiency [@stephens1986], but other work suggests that both generalist and specialist strategies could be adaptive as resources fluctuate [@woo.etal2008;@phillips.etal2017]. Our results support empirical findings indicating that flexible and specialist foragers obtained similar success given prey behaviour [@weimerskirch2007;@woo.etal2008;@potier.etal2015;@phillips.etal2017;@courbin.etal2018], although generalist hunters were slightly less successful at more advanced stages. Both strategies appeared to emerge in part from individuals learning how to hunt their prey, resulting in a general increase in success in the population. However, there were still considerable differences in success among individuals through time, suggesting that some predators were limited in their capacity to match their tactic to their prey or to increase their success through other means.

Specialist foragers were faster and probably better equipped to hunt the faster and more difficult prey in *DBD* [@fraserfranco.etal2024]. However, if the prey responded to fast predators by also being faster, then hunting at high speeds resulted in challenging encounters for these predators too, thereby decreasing the benefits of using this tactic (Figure 4). Thus, specialising probably compensated for the difficulty of hunting prey at high speeds by helping predators to better predict the location and movement of their prey. On the other hand, generalist hunters encountered a larger range of prey moving at slower speeds on average. Yet, because the prey increased their speed with experience, the benefits of being able to hunt multiple prey types for generalist hunters may have come at the cost of not being adept at capturing faster prey [@pintor.etal2014]. Such trade-offs may reflect limitations in learning all the skills required to successfully hunt all types of prey [@healy1992;@belisle1997;@dukas2019], particularly if the skills required to hunt slower prey are nontransferable to faster prey. As the benefits of each hunting style changed under different scenarios, the combination of predatory styles in this virtual system reflects how resource fluctuations can maintain fitness equilibrium within populations, resulting in behavioural diversity in predator-prey systems [@tinker.etal2008;@woo.etal2008;@mougi.kishida2009].

@@

We found support for our prediction that individual foraging specialisation changed across experience and predator-prey interactions. Our results suggest that predators learned with experience, as their success increased and their speed remained matched to the speed of their prey.  Even though individuals were not all equally successful, both specialised and generalist hunters achieved similar success overall. A potential caveat is that generalist hunters might have experimented with various tactics out of boredom, which could impede ecologically realistic interpretations of our data. However, the consistent association between this tactic and heterogeneous prey groups gives us confidence that it emerged from their interactions. One limitation of our study was that we couldn't monitor all the matches of the prey, which prevented us from assessing their responses to the predator through their experience. Hence, future studies should aim at monitoring reciprocal behavioural dynamics through time, which may reveal important insights on the mechanisms driving predator-prey systems.
